{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-04-29 paper.tex
hspace for figure
adding split plot
more edits

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard-scattering cross
section. Consider for example the calculation of a general cross
section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation} \label{eq:hadconv}
  \sigma_{pp\to X} =
  \sum_{i,j}^{N_F}\int dx_1\, dx_2\, f_i(x_1,Q^2)\;
  \hat{\sigma}_{ij\to X}\; f_j(x_2,Q^2)\,,
\end{equation}
where $Q^2$ is the typical hard scale of the process and the indices
$i,j$ sum over the active partonic species in the calculation. The
partonic cross section $\hat{\sigma}_{ij\to X}$ admits the
perturbative expansion
\begin{equation}\label{eq:expansion}
\hat{\sigma}_{ij\to X} = \alpha_s^{p_{\rm LO}}(Q^2) \sum_{p=0}^{N_\alpha} \alpha_s^{p}(Q^2)\hat{\sigma}^{(p)}_{ij\to X}
\end{equation}
$p_{\rm LO}$ being the $\alpha_s$ power of the leading-order
contribution and $\hat{\sigma}^{(p)}_{ij\to X}$ the N$^p$LO
contribution. Therefore to order $N_\alpha$ in QCD
\begin{equation}\label{eq:pertexp}
  \sigma_{pp\to X} =
  \sum_{i,j}^{N_F}\sum_{p=0}^{N_\alpha} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)}_{ij\to X}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)\,.
\end{equation}

The central observation of tools such as {\tt APPLgrid} and {\tt
  FastNLO} is that the PDF and $\alpha_S$ dependence may be factorised out of
  the convolution via expansion over a set of interpolating functions in $Q^2$ and the
  two values of parton-$x$. Explicitly, one may represent the PDFs and $\alpha_S$ as
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)
  \,\mathcal{I}_\tau^{(n')}(Q^2)\,\mathcal{I}_\alpha^{(n_1)}(x_1)
  \,\mathcal{I}_\beta^{(n_2)}(x_2)
\end{array}
\end{equation}
in terms of interpolating functions $\mathcal{I}_\tau^{(n')}(Q^2)$,
$\mathcal{I}_\alpha^{(n_1)}(x_1)$, $\mathcal{I}_\beta^{(n_2)}(x_2)$
(typically Lagrange polynomials) of degree $n'$, $n_1$, and $n_2$.
Using these expressions for the PDFs in the double-convolution of Eq.~(\ref{eq:pertexp}) one finds:
\begin{equation}\label{eq:XsecPDFs}
\begin{array}{rcl}
  \displaystyle \sigma_{pp\to X} &=& \displaystyle 
                                     \sum_{i,j}\sum_{p=0}^{N_\alpha} \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
                                     \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\\
  \\
                                 &\times& \displaystyle \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
                                          \hat{\sigma}^{(p)}_{ij\to X}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
                                          \,\mathcal{I}_\beta^{(n_2)}(x_2)\,.
\end{array}
\end{equation}

Additionally one may take advantages of symmetries present in the matrix element to reduce the sum over active flavours $(i,j)$ in
Eq.~(\ref{eq:hadconv}) to a reduced number of partonic subprocesses with a replacement such as
\begin{equation}
\sum_{i,j} \hat{\sigma}^{(p)}_{ij\to X}
f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\rightarrow
\sum_{s}^{N_{\rm sub}} \hat{\sigma}^{(p)(s)} F^{(s)}_{\alpha\beta,\tau}
\end{equation}
where the $s$ index runs over the $N_{\rm sub}$ (process-specific)
parton subprocess:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}_{\alpha\beta,\tau} =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{\alpha},Q^2_\tau)f_j(x_{\beta},Q^2_\tau)\,,
\end{equation}
where the $C^{(s)}_{ij}$ matrix encodes the combination of PDFs that
contribute to the $s$-th subprocess and $\hat{\sigma}^{(p)(s)}$
is the corresponding hard cross section. Finally, defining:
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
  \,\mathcal{I}_\beta^{(n_2)}(x_2)\,,
\end{equation}
one can write Eq.~(\ref{eq:XsecPDFs}) as follows:
\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}

@@

\section{{\tt APFELgrid} manual}

\section{Benchmark}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ph]
\centering
\includegraphics[scale=0.5]{plots/t0}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering cross-section. Consider for example the calculation of a general cross-section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation}\label{eq:pertexp}
  \sigma_{pp\to X} =
  \sum_{i,j}\sum_{p=0} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)}_{ij\to X}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)\,.
\end{equation}
where $Q^2$ is the typical hard scale of the process, the indices
$i,j$ sum over the active partonic species in the calculation, $p_{\rm LO}$ is the leading-order power of $\alpha_s$ for the process
and $\hat{\sigma}^{(p)}_{ij\to X}$ the N$^p$LO contribution.

The central observation of tools such as {\tt APPLgrid} and {\tt
  FastNLO} is that the PDF and $\alpha_S$ dependence may be factorised out of
  the convolution via expansion over a set of interpolating functions, spanning $Q^2$ and the
  two values of parton-$x$. For example one may represent the PDFs and $\alpha_S$ as
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)
  \,\mathcal{I}_\tau(Q^2)\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)
\end{array}
\end{equation}
in terms of the Lagrange basis polynomials $\mathcal{I}_\tau(Q^2)$,
$\mathcal{I}_\alpha(x_1)$, $\mathcal{I}_\beta(x_2)$.
Using these expressions for the PDFs in the double-convolution of Eq.~(\ref{eq:pertexp}) one finds:
\begin{equation}\label{eq:XsecPDFs}
\begin{array}{rcl}
  \displaystyle \sigma_{pp\to X} &=& \displaystyle 
                                     \sum_{i,j}\sum_{p=0}^{N_\alpha} \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
                                     \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\\
  \\
                                 &\times& \displaystyle \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
                                          \hat{\sigma}^{(p)}_{ij\to X}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
                                          \,\mathcal{I}_\beta^{(n_2)}(x_2)\,.
\end{array}
\end{equation}

Additionally one may take advantages of symmetries present in the matrix element to reduce the sum over active flavours $(i,j)$ in
Eq.~(\ref{eq:hadconv}) to a reduced number of partonic subprocesses with a replacement such as
\begin{equation}
\sum_{i,j} \hat{\sigma}^{(p)}_{ij\to X}
f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\rightarrow
\sum_{s}^{N_{\rm sub}} \hat{\sigma}^{(p)(s)} F^{(s)}_{\alpha\beta,\tau}
\end{equation}
where the $s$ index runs over the $N_{\rm sub}$ (process-specific)
parton subprocess:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}_{\alpha\beta,\tau} =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{\alpha},Q^2_\tau)f_j(x_{\beta},Q^2_\tau)\,,
\end{equation}
and the $C^{(s)}_{ij}$ matrix encodes the combination of PDFs that
contribute to the $s$-th subprocess and $\hat{\sigma}^{(p)(s)}$
is the corresponding subprocess hard cross section. Finally, defining:
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
  \,\mathcal{I}_\beta^{(n_2)}(x_2)\,,
\end{equation}
one can write Eq.~(\ref{eq:XsecPDFs}) as follows:
\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}

@@

\section{Performance benchmarks}

@@

\section{Conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ph]
  \centering
  \hspace*{-2.0cm}                                                           
  \includegraphics[scale=0.5]{plots/t0}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ph]
  \centering
  \includegraphics[scale=0.5]{plots/t0a}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ph]
  \centering                                      
  \includegraphics[scale=0.5]{plots/t0b}
\caption{\small Performance comparisons between {\tt APFELgrid} with
  AVX, SSE3 and double precision convolution and {\tt APPLgrid}
  convolution time per point and process.}
\label{fig:benchmark}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
